|

### 2023-05-03 paper.Rmd
rework 2.3

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland1984graphical]. Human examination of diagnostic plots can feel subjective. Ideally we would expect visual evaluation to be done objectively with small amounts of human-to-human variation in reading data plots, but we cannot assume this is the case.

In practice, over-interpretation of a single plot is common.  For instance, @roy2015using described a published example where authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis. @roy2015using showed that there were no differences in the expression levels between the gene groups. One solution to over-interpretation is to examine the plot in the context of natural sampling variability assumed by the model, called the lineup protocol, as proposed in @buja2009statistical.  @majumder2013validation showed that the lineup protocol is analogous to the null hypothesis significance testing framework. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, are constructed using the same graphical procedure as the data plot but the data is replaced with null data that is generated in a manner consistent with the null hypothesis, $H_0$. Then, an observer who has not seen the data plot is asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires a priori specification of $H_0$ (or at least a null data generating mechanism), much like the requirement of knowing the sampling distribution of the test statistic in null hypothesis significance testing framework.

@@

Data used in the $m - 1$ null plots needs to be simulated. In regression diagnostics, sampling data consistent with $H_0$ is equivalent to sampling data from the assumed model. As @buja2009statistical suggested, $H_0$ is usually a composite hypothesis controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to a so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja2009statistical]. Essentially, null residuals can be simulated by regressing $N$ i.i.d standard normal random draws on the predictors, then re-scaling it by the ratio of residual sum of square in two regressions.

The effectiveness of lineup protocol for regression analysis has been validated by @majumder2013validation under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power to a t-test, though they express that in general it is unnecessary to use visual inference if there exists a conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where there is not a conventional test, visual test outperforms the conventional test by a large margin. This is encouraging, as it supports the use of visual inference in situations where there are no existing statistical testing procedures. Visual inference has also been integrated into diagnostics for hierarchical linear models where the lineup protocol is used to judge the assumptions of linearity, normality and constant error variance for both the level-1 and level-2 residuals [@loy2013diagnostic, @loy2014hlmdiag and @loy2015you].

>>>>

The examination of data plots to infer signal or patterns (or lack thereof) is fraught with variation in the human ability to interpret and decode the information embedded in a graph [@cleveland1984graphical]. Human examination of diagnostic plots can feel subjective. Ideally we would expect visual evaluation to be done objectively with low human-to-human variation in reading data plots, but we cannot assume this is the case.

In practice, over-interpretation of a single plot is common.  For instance, @roy2015using described a published example where authors over-interpreted separation between gene groups from a two-dimensional projection of a linear discriminant analysis even when there were no differences in the expression levels between the gene groups. One solution to over-interpretation is to examine the plot in the context of natural sampling variability assumed by the model, called the lineup protocol, as proposed in @buja2009statistical.  @majumder2013validation showed that the lineup protocol is analogous to the null hypothesis significance testing framework. The protocol consists of $m$ randomly placed plots, where one plot is the data plot, and the remaining $m - 1$ plots, referred to as the _null plots_, are constructed using the same graphical procedure as the data plot but the data is replaced with null data that is generated in a manner consistent with the null hypothesis, $H_0$. Then, an observer who has not seen the data plot is asked to point out the most different plot from the lineup. Under $H_0$, it is expected that the data plot would have no distinguishable difference from the null plots, and the probability that the observer correctly picks the data plot is $1/m$. If one rejects $H_0$ as the observer correctly picks the data plot, then the Type I error of this test is $1/m$. This protocol requires a priori specification of $H_0$ (or at least a null data generating mechanism), much like the requirement of knowing the sampling distribution of the test statistic in null hypothesis significance testing framework.

@@

Data used in the $m - 1$ null plots needs to be simulated. In regression diagnostics, sampling data consistent with $H_0$ is equivalent to sampling data from the assumed model. As @buja2009statistical suggested, $H_0$ is usually a composite hypothesis controlled by nuisance parameters. Since regression models can have various forms, there is no general solution to this problem, but it sometimes can be reduced to a so called "reference distribution" by applying one of the three methods: (i) sampling from a conditional distribution given a minimal sufficient statistic under $H_0$, (ii) parametric bootstrap sampling with nuisance parameters estimated under $H_0$, and (iii) Bayesian posterior predictive sampling. The conditional distribution given a minimal sufficient statistic is the best justified reference distribution among the three [@buja2009statistical]. Under this method, the null residuals can essentially be simulated by independent drawing from  a standard normal random distribution, then regressing the draws on the predictors, and then re-scaling it by the ratio of the residual sum of square in two regressions.

The effectiveness of lineup protocol for regression analysis has been validated by @majumder2013validation under relatively simple settings with up to two predictors. Their results suggest that visual tests are capable of testing the significance of a single predictor with a similar power to a t-test, though they express that in general it is unnecessary to use visual inference if there exists a corresponding conventional test, and they do not expect the visual test to perform equally well as the conventional test. In their third experiment, where the contamination of the data violate the assumptions of the convetional test, visual test outperforms the conventional test by a large margin. This is encouraging, as it supports the use of visual inference in situations where there are no existing numerical testing procedures. Visual inference has also been integrated into diagnostics for hierarchical linear models where the lineup protocol is used to judge the assumptions of linearity, normality and constant error variance for both the level-1 and level-2 residuals [@loy2013diagnostic, @loy2014hlmdiag and @loy2015you].
